>

### 2020-08-24 paper/spyrison-cook.Rmd
caching issue, view_basis to oblique_frame

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.
keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book ## alternative
---
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center", 
  echo       = TRUE, 
  collapse   = TRUE,
  message    = FALSE, 
  warning    = FALSE, 
  error      = FALSE,
  cache      = TRUE, # needed for log and figure out?
  cache.lazy = FALSE
)
if (F){ ## From reproducing, please use this dev ver:
  remotes::install_github("nspyrison/spinifex", 
                          ref = "da75976588dc008e21f01ce6a57d9778d0ed6933")
}
library(tourr)
library(spinifex) ## Use dev version listed above!
library(ggplot2)
library(gridExtra)
library(dplyr)
```

@@

view_manip_space2 <- function (basis,
                               manip_var,
                               manip_col = "blue",
                               tilt = pi * 5 / 12,
                               z_col = "red",
                               labels = paste0("V", 1:nrow(basis)))
{
  #### Finds the angle between two vectors
  find_angle <- function(a, b)
    acos(sum(a * b) / (sqrt(sum(a * a)) * sqrt(sum(b * b))) )
  #### Makes a df semi-circle, with 1 row per degree
  make_curve <- function(ang_st = 0,
                         ang_stop = 2 * pi) {
      degrees <- round(360 / (2 * pi) * abs(ang_st - ang_stop))
      angle <- seq(ang_st, ang_stop, length = degrees)
      data.frame(x = cos(angle), y = sin(angle), z = sin(angle))
    }

@@

  ## Render
  gg <-
    ggplot2::ggplot() + ggplot2::scale_color_brewer(palette = "Dark2") +
    ggplot2::theme_void() + ggplot2::theme(legend.position = "none") +
    ggplot2::coord_fixed() +
    ## Circle
    ggplot2::geom_path(data = circ_r,
                       mapping = ggplot2::aes(x = x, y = y),
                       color = manip_col, size = 0.3, inherit.aes = FALSE) +
    ## Projection plane:
    ggplot2::geom_segment(data = m_sp_r,
                          mapping = ggplot2::aes(
                            x = x,
                            y = y,
                            xend = 0,
                            yend = 0
                          ),
                          size = siz_v, colour = col_v) +
    ggplot2::geom_text(data = m_sp_r,
                       mapping = ggplot2::aes(x = x, y = y, label = labels),
                       size = 4, colour = col_v,
                       vjust = "outward", hjust = "outward") +
    ## Z direction
    ggplot2::geom_path(data = circ_r,
                       mapping = ggplot2::aes(x = x, y = z),
                       color = z_col, size = 0.3, inherit.aes = FALSE) +
    ggplot2::geom_segment(data = mvar_r,
                          mapping = ggplot2::aes(x = x,y = z,
                                                 xend = 0,yend = 0),
                          size = 1, colour = z_col) +
    ggplot2::geom_segment(data = mvar_r,
                          mapping = ggplot2::aes(x = x, y = z,
                                                 xend = x,yend = y),
                          size = 0.3, colour = "grey80", linetype = 2) +
    ggplot2::geom_text(data = mvar_r,
                       mapping = ggplot2::aes(x = x, y = z, 
                                              label = labels[manip_var]),
                       size = 4, colour = z_col,
                       vjust = "outward",
                       hjust = "outward"
    ) +
    ## Label angles
    geom_path(data = phi_curve_r,
              mapping = aes(x=x, y=z),
              color = z_col, size = 0.2) +
    geom_text(data = 1.2 * phi_curve_r[ceiling(nrow(phi_curve_r)/2), ],
              mapping = ggplot2::aes(x=x, y=z, label = "phi"),
              color = z_col, size = 4, parse = TRUE) +
    geom_path(data = theta_curve_r,
              mapping = aes(x, y),
              color = manip_col,
              size = 0.2) +
    geom_text(data = 1.2 * theta_curve_r[ceiling(nrow(theta_curve_r)/2), ],
              mapping = ggplot2::aes(x=x, y=y-.02, label = "theta"),
              color = z_col, size = 4, parse = TRUE)

@@

The function `view_basis` can be used to show a projection of the basis, or with the data overlaid. For example, the plots in Figures \ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

>>>>

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.
keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book ## alternative
---
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center", 
  echo       = TRUE, 
  collapse   = TRUE,
  message    = FALSE, 
  warning    = FALSE, 
  error      = FALSE,
  cache      = FALSE, # needed for log and figure out?
  cache.lazy = FALSE
)
if (F){ ## From reproducing, please use this dev ver:
  remotes::install_github("nspyrison/spinifex", 
                          ref = "da75976588dc008e21f01ce6a57d9778d0ed6933")
}
library(tourr)
library(spinifex) ## Use dev version listed above!
library(ggplot2)
library(gridExtra)
library(dplyr)
```

@@

view_manip_space2 <- function (basis,
                               manip_var,
                               manip_col = "blue",
                               tilt = pi * 5 / 12,
                               z_col = "red",
                               lab = paste0("V", 1:nrow(basis)))
{
  #### Finds the angle between two vectors
  find_angle <- function(a, b)
    acos(sum(a * b) / (sqrt(sum(a * a)) * sqrt(sum(b * b))) )
  #### Makes a df semi-circle, with 1 row per degree
  make_curve <- function(ang_st = 0,
                         ang_stop = 2 * pi) {
      degrees <- round(360 / (2 * pi) * abs(ang_st - ang_stop))
      angle <- seq(ang_st, ang_stop, length = degrees)
      data.frame(x = cos(angle), y = sin(angle), z = sin(angle))
    }

@@

  ## Render
  gg <-
    ggplot2::ggplot() + ggplot2::scale_color_brewer(palette = "Dark2") +
    ggplot2::theme_void() + ggplot2::theme(legend.position = "none") +
    ggplot2::coord_fixed() +
    ## Circle
    ggplot2::geom_path(data = circ_r,
                       mapping = ggplot2::aes(x = x, y = y),
                       color = manip_col, size = 0.3, inherit.aes = FALSE) +
    ## Projection plane:
    ggplot2::geom_segment(data = m_sp_r,
                          mapping = ggplot2::aes(
                            x = x,
                            y = y,
                            xend = 0,
                            yend = 0
                          ),
                          size = siz_v, colour = col_v) +
    ggplot2::geom_text(data = m_sp_r,
                       mapping = ggplot2::aes(x = x, y = y, label = lab),
                       size = 4, colour = col_v,
                       vjust = "outward", hjust = "outward") +
    ## Z direction
    ggplot2::geom_path(data = circ_r,
                       mapping = ggplot2::aes(x = x, y = z),
                       color = z_col, size = 0.3, inherit.aes = FALSE) +
    ggplot2::geom_segment(data = mvar_r,
                          mapping = ggplot2::aes(x = x,y = z,
                                                 xend = 0,yend = 0),
                          size = 1, colour = z_col) +
    ggplot2::geom_segment(data = mvar_r,
                          mapping = ggplot2::aes(x = x, y = z,
                                                 xend = x,yend = y),
                          size = 0.3, colour = "grey80", linetype = 2) +
    ggplot2::geom_text(data = mvar_r,
                       mapping = ggplot2::aes(x = x, y = z, 
                                              label = lab[manip_var]),
                       size = 4, colour = z_col,
                       vjust = "outward",
                       hjust = "outward"
    ) +
    ## Label angles
    geom_path(data = phi_curve_r,
              mapping = aes(x = x, y = z),
              color = z_col, size = 0.2) +
    geom_text(data = 1.2 * phi_curve_r[ceiling(nrow(phi_curve_r) / 2), ],
              mapping = ggplot2::aes(x=x, y=z, label = "phi"),
              color = z_col, size = 4, parse = TRUE) +
    geom_path(data = theta_curve_r,
              mapping = aes(x, y),
              color = manip_col,
              size = 0.2) +
    geom_text(data = 1.2 * theta_curve_r[ceiling(nrow(theta_curve_r)/2), ],
              mapping = ggplot2::aes(x = x, y = y - .02, label = "theta"),
              color = manip_col, size = 4, parse = TRUE)

@@

The function `oblique_frame` can be used to show a projection of the basis, or with the data overlaid. For example, the plots in Figures \ref{fig:step0} and \ref{fig:step3} were made with code similar to this:
